A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-04-14 manuscript.qmd
switch to quarto: - add manuscript.qmd and _quarto.yml - add apaquarto extension and apa.csl - update .gitignore and move results.rds to main project directory

---
title             : "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle        : "Simulation-Based Sample Size Planning for GLMMs"
author:

  - name          : "Timo K. Koch"
    affiliation   : "1, 2"
    corresponding : yes    # Define only one corresponding author
    address       : "Torstrasse 25, 9000 St. Gallen, Switzerland"
    email         : "timo.koch@unisg.ch"
    equal_contrib: yes
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Formal Analysis"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Florian Pargent"
    affiliation   : "1"
    equal_contrib: yes
    role:
      - "Formal Analysis"
      - "Methodology"
      - "Visualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Anne-Kathrin Kleine"
    affiliation   : "1"
    role:
      - "Formal Analysis"      
      - "Writing - Review & Editing"
  - name          : "Eva Lermer"
    affiliation   : "1,3"
    role:
      - "Conceptualization"
      - "Resources"
      - "Writing - Review & Editing"
  - name          : "Susanne Gaube"
    affiliation   : "1,4"
    role:
      - "Conceptualization"
      - "Supervision"
      - "Writing - Review & Editing"
affiliation:
  - id            : "1"
    institution   : "LMU Munich, Department of Psychology"
  - id            : "2"
    institution   : "University of St. Gallen, Institute of Behavioral Science & Technology"
  - id            : "3"
    institution   : "Technical University of Applied Sciences Augsburg, Department of Business Psychology"
  - id            : "4"
    institution   : "University College London, Global Business School for Health"
note: "\\clearpage" #added this to move entire author note to second page
authornote: |
  This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525).
  All materials (reproducible manuscript incl. R code, simulation results) are available in the project's repository on the Open Science Framework (OSF): https://osf.io/dhwf4/
abstract: |
  When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, available methods and software for a priori power analyses for GLMMs are often limited to specific designs, while data simulation approaches offer more flexibility. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research.

keywords          : "power analysis, data simulation, sample size, generalized linear mixed model"
wordcount         : "7504"
bibliography      : ["../Simulation Tutorial Paper.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output: 
  papaja::apa6_pdf:
---

```{r setup, include = FALSE}
library("papaja")
library("tidyverse")
library("faux")
library("lme4")
library("multcomp")
library("marginaleffects")
library("future")
library("furrr")
library("binom")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, echo = TRUE, fig.keep = "high", dev = "jpeg")
```

@@

```{r, echo = FALSE}
b_0 = 0.847
b_e = 1.350
b_a = -1.253
b_c = 2.603
b_ea = 0.790
b_ec = -1.393
```

@@

```{r probtable, echo=FALSE}
probtable <- data.frame(
  col1 = 
    c("no advice, no expert", "no advice, expert", "false advice, no expert", 
      "false advice, expert", "correct advice, no expert", "correct advice, expert"),
  col2 =
    c("0.70", "0.90", "0.40", "0.85", "0.90", "0.95"),
  col3 = 
    c("$logit(0.70) = \\beta_0$",
      "$logit(0.90) = \\beta_0 + \\beta_e$", 
      "$logit(0.40) = \\beta_0 + \\beta_a$",
      "$logit(0.85) = \\beta_0 + \\beta_e + \\beta_{a} + \\beta_{ea}$",
      "$logit(0.90) = \\beta_0 + \\beta_a + \\beta_c$",
      "$logit(0.95) = \\beta_0 + \\beta_e + \\beta_a + \\beta_c + \\beta_{ea} + \\beta_{ec}$"))
apa_table(
  probtable
  , col.names = c("Experimental condition",
    "$P(Y=1|\\mathbf{X=x}, \\mathbf{U} = \\mathbf{0})$",
    "Implied equation")
  , caption = "Assumed probabilities that an average subject solves an average item in each experimental condition."
  , note = "Implied equations are derived based on the model equations and setting all random intercept terms to 0."
  , escape = FALSE
)
```

@@

(ref:margdist1-caption) Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions.

```{r margdist1, fig.cap="(ref:margdist1-caption)"}
library(ggdist)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

Figure \@ref(fig:margdist1) shows the model implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered in the previous section. This difference increases with the variability of the random effects.

@@

(ref:margdist2-caption) Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of item random intercepts to 0.01.

```{r margdist2, echo=FALSE, fig.cap="(ref:margdist2-caption)"}
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0i = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

As an example, Figure \@ref(fig:margdist2) reveals a number of implicit assumptions about the comparison between experts and non-experts: With wrong advice, virtually all experts have a higher probability of making a correct diagnosis compared to non-experts when considering only items with average difficulty. In contrast, there is considerable overlap in probability between experts and non-experts with no advice and even higher overlap with correct advice. Patterns like these should be considered carefully and discussed with the domain experts. Parameter values ($\beta$ parameters, and $\sigma_S$) should be adjusted if the implications do not seem reasonable.

@@

```{r, include=FALSE}
set.seed(1)
dat <- simulate(n_subjects = 2000, n_items = 2000, sd_u0s = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For Figure \@ref(fig:margdist3), we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. However, these high standard deviations do not account for the inherent variability and complexity of human performance. For example, we would expect that a participant with low ability compared to other task experts to solve a difficult item with a probability substantially larger than zero even when presented with wrong advice.

(ref:margdist3-caption) Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of subject and item random intercepts to 3.

```{r margdist3, echo=FALSE, fig.cap="(ref:margdist3-caption)"}
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 3, sd_u0i = 3)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

```{r, message=FALSE}
library(tidyverse)
set.seed(1)
dat <- simulate(n_subjects = 100, n_items = 50)
```

@@

```{r, message=FALSE}
library(lme4)
f <- y_bin ~ 1 + expert + advice_present + advice_correct + 
  expert:advice_present + expert:advice_correct +
  (1|subject) + (1|item)
```

@@

```{r, message=FALSE, eval=FALSE}
library(future)
plan("multisession", workers = 6)
set.seed(2)
```

@@

```{r, message=FALSE, echo=FALSE, eval=FALSE}
library(future)
library(furrr)
plan("multisession", workers = 6)
set.seed(2)
sim_result <- crossing(
  rep = 1:300,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) %>%
  mutate(res = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) %>%
  unnest(col = res)
saveRDS(sim_result, file = "results.rds")
```

```{r, message=FALSE, eval=FALSE}
library(furrr)
sim_result <- crossing(
  rep = 1:300,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) %>%
  mutate(res = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) %>%
  unnest(col = res)
```

@@

```{r, echo=FALSE}
sim_result <- readRDS(file = "results.rds")
```

```{r finalpwr, fig.cap="Simulation-based power estimates including 95% confidence interval of the case study for different numbers of subjects and items, based on a significance level of 0.05."}
library(binom)
alpha <- 0.05
power <- sim_result %>%
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) %>%
  group_by(n_subjects, n_items) %>% 
  summarise(
    power = mean(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha), 
    n_sig = sum(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  xlab("number of subjects") + ylab("number of items")
```

@@

As should be the case, power estimates in Figure \@ref(fig:finalpwr) increase with both the number of subjects and the number of items. The confidence intervals reported here indicate how precisely power was estimated by our simulation. Higher precision (which would be reflected in narrower confidence intervals) could be obtained by increasing the number of repetitions (`rep`) in the simulation. In practice, data simulations are often run multiple times with adjusted combinations of sample sizes. When running for the first time, it might be revealed that power is way too low (or much higher than required) for some combinations of `n_subjects` and `n_items`. When narrowing down the best combination that achieves sufficient power while at the same time striking a good balance of how many subjects and items are practically feasible, later rounds of data simulation will typically include a smaller grid of sample sizes combined with a higher number of repetitions. This will assure high precision for the final power estimates, which are then used for the sample size justification of the future study.

@@

```{r finalprecision, fig.cap="Simulation-based precision estimates (expected width of confidence intervals) including 95% confidence interval of the case study for different numbers of subjects and items, based on a confidence level of 0.95."}
precision <- sim_result %>%
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) %>%
  group_by(n_subjects, n_items) %>%
  mutate(width = `conf.high_b8=b2` - `conf.low_b8=b2`) %>%
  summarise(precision = mean(width),
    ci.lwr = t.test(width)$conf.int[1],
    ci.upr = t.test(width)$conf.int[2], 
    .groups = "drop")
precision %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = precision)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                precision, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 0.3), direction = -1) +
  guides(fill = guide_legend(reverse=FALSE))
```

As should be the case, precision estimates in Figure \@ref(fig:finalprecision) increase  (i.e., average width of confidence interval decreases) with both the number of subjects and the number of items.
The confidence intervals reported here indicate how precisely the expected width of the confidence interval for our focal contrast was estimated by our simulation.
Applying our simple heuristic of targeting an expected width smaller than 0.1, we see the same trade-off between the number of subjects and the number of items as with planning for power.
We could either choose 100 subjects and 30 items or 200 subjects and 10 items.

@@

\newpage

@@

::: {#refs custom-style="Bibliography"}
:::

>>>>

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
author:
  - name: Timo K. Koch
    corresponding: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Florian Pargent
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - ref: lmu
  - name: Anne-Kathrin Kleine
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    roles:
      - conceptualization
      - funding aquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: Global Business School for Health
author-note:
  disclosures:
    data-sharing: "All materials (reproducible manuscript incl. R code, simulation results) are available in the project's repository on the Open Science Framework (OSF) at https://osf.io/dhwf4/"
    conflict-of-interest: "None."
    financial-support: "This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525)."
    gratitude: ~
    authorship-agreements: "Timo K. Koch and Florian Pargent contributed equally to this work."
abstract: "When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, available methods and software for a priori power analyses for GLMMs are often limited to specific designs, while data simulation approaches offer more flexibility. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research."
keywords: [power analysis, data simulation, sample size, generalized linear mixed model]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

```{r}
#| label: setup
#| include: false
library("papaja")
library("tidyverse")
library("faux")
library("lme4")
library("multcomp")
library("marginaleffects")
library("future")
library("furrr")
library("binom")
r_refs("r-references.bib")
```

@@

```{r}
#| echo: false
b_0 = 0.847
b_e = 1.350
b_a = -1.253
b_c = 2.603
b_ea = 0.790
b_ec = -1.393
```

@@

```{r}
#| label: probtable
#| echo: false
probtable <- data.frame(
  col1 = 
    c("no advice, no expert", "no advice, expert", "false advice, no expert", 
      "false advice, expert", "correct advice, no expert", "correct advice, expert"),
  col2 =
    c("0.70", "0.90", "0.40", "0.85", "0.90", "0.95"),
  col3 = 
    c("$logit(0.70) = \\beta_0$",
      "$logit(0.90) = \\beta_0 + \\beta_e$", 
      "$logit(0.40) = \\beta_0 + \\beta_a$",
      "$logit(0.85) = \\beta_0 + \\beta_e + \\beta_{a} + \\beta_{ea}$",
      "$logit(0.90) = \\beta_0 + \\beta_a + \\beta_c$",
      "$logit(0.95) = \\beta_0 + \\beta_e + \\beta_a + \\beta_c + \\beta_{ea} + \\beta_{ec}$"))
apa_table(
  probtable
  , col.names = c("Experimental condition",
    "$P(Y=1|\\mathbf{X=x}, \\mathbf{U} = \\mathbf{0})$",
    "Implied equation")
  , caption = "Assumed probabilities that an average subject solves an average item in each experimental condition."
  , note = "Implied equations are derived based on the model equations and setting all random intercept terms to 0."
  , escape = FALSE
)
```

@@

```{r}
#| label: fig-margdist1
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions."

library(ggdist)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@fig-margdist1 shows the model implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered in the previous section. This difference increases with the variability of the random effects.

@@

```{r}
#| label: fig-margdist2
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of item random intercepts to 0.01."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0i = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

As an example, @fig-margdist2 reveals a number of implicit assumptions about the comparison between experts and non-experts: With wrong advice, virtually all experts have a higher probability of making a correct diagnosis compared to non-experts when considering only items with average difficulty. In contrast, there is considerable overlap in probability between experts and non-experts with no advice and even higher overlap with correct advice. Patterns like these should be considered carefully and discussed with the domain experts. Parameter values ($\beta$ parameters, and $\sigma_S$) should be adjusted if the implications do not seem reasonable.

@@

```{r}
#| include: false
set.seed(1)
dat <- simulate(n_subjects = 2000, n_items = 2000, sd_u0s = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For @fig-margdist3 we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. However, these high standard deviations do not account for the inherent variability and complexity of human performance. For example, we would expect that a participant with low ability compared to other task experts to solve a difficult item with a probability substantially larger than zero even when presented with wrong advice.

```{r}
#| label: fig-margdist3
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of subject and item random intercepts to 3."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 3, sd_u0i = 3)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

```{r}
#| message: false
library(tidyverse)
set.seed(1)
dat <- simulate(n_subjects = 100, n_items = 50)
```

@@

```{r}
#| message: false
library(lme4)
f <- y_bin ~ 1 + expert + advice_present + advice_correct + 
  expert:advice_present + expert:advice_correct +
  (1|subject) + (1|item)
```

@@

```{r}
#| message: false
#| eval: false
library(future)
plan("multisession", workers = 6)
set.seed(2)
```

@@

```{r}
#| echo: false
#| eval: false
library(future)
library(furrr)
plan("multisession", workers = 6)
set.seed(2)
sim_result <- crossing(
  rep = 1:300,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) %>%
  mutate(res = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) %>%
  unnest(col = res)
saveRDS(sim_result, file = "results.rds")
```

```{r}
#| message: false
#| eval: false
library(furrr)
sim_result <- crossing(
  rep = 1:300,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) %>%
  mutate(res = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) %>%
  unnest(col = res)
```

@@

```{r}
#| echo: false
sim_result <- readRDS(file = "results.rds")
```

```{r}
#| label: fig-finalpwr
#| fig-cap: "Simulation-based power estimates including 95% confidence interval of the case study for different numbers of subjects and items, based on a significance level of 0.05."
library(binom)
alpha <- 0.05
power <- sim_result %>%
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) %>%
  group_by(n_subjects, n_items) %>% 
  summarise(
    power = mean(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha), 
    n_sig = sum(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  xlab("number of subjects") + ylab("number of items")
```

@@

As should be the case, power estimates in @fig-finalpwr increase with both the number of subjects and the number of items. The confidence intervals reported here indicate how precisely power was estimated by our simulation. Higher precision (which would be reflected in narrower confidence intervals) could be obtained by increasing the number of repetitions (`rep`) in the simulation. In practice, data simulations are often run multiple times with adjusted combinations of sample sizes. When running for the first time, it might be revealed that power is way too low (or much higher than required) for some combinations of `n_subjects` and `n_items`. When narrowing down the best combination that achieves sufficient power while at the same time striking a good balance of how many subjects and items are practically feasible, later rounds of data simulation will typically include a smaller grid of sample sizes combined with a higher number of repetitions. This will assure high precision for the final power estimates, which are then used for the sample size justification of the future study.

@@

```{r}
#| label: fig-finalprecision
#| fig-cap: "Simulation-based precision estimates (expected width of confidence intervals) including 95% confidence interval of the case study for different numbers of subjects and items, based on a confidence level of 0.95."
precision <- sim_result %>%
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) %>%
  group_by(n_subjects, n_items) %>%
  mutate(width = `conf.high_b8=b2` - `conf.low_b8=b2`) %>%
  summarise(precision = mean(width),
    ci.lwr = t.test(width)$conf.int[1],
    ci.upr = t.test(width)$conf.int[2], 
    .groups = "drop")
precision %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = precision)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                precision, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 0.3), direction = -1) +
  guides(fill = guide_legend(reverse=FALSE))
```

As should be the case, precision estimates in @fig-finalprecision increase  (i.e., average width of confidence interval decreases) with both the number of subjects and the number of items.
The confidence intervals reported here indicate how precisely the expected width of the confidence interval for our focal contrast was estimated by our simulation.
Applying our simple heuristic of targeting an expected width smaller than 0.1, we see the same trade-off between the number of subjects and the number of items as with planning for power.
We could either choose 100 subjects and 30 items or 200 subjects and 10 items.

@@

@@

::: {#refs}
:::
